\sysfull

### 2018-04-17 paper/background.tex
edits to background

%Here, we provide background on energy harvesting systems and the intermittent software execution model. We then discuss the key limitations of existing task-based execution models for intermittent computing that \sys addresses.

%\begin{figure}
\begin{wrapfigure}{t!}{0.5\textwidth}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example}
		\caption{Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration.}\label{fig:crc_example}
	\end{subfigure}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example_war}
		\caption{Execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 9) with stale reminder (line 3)}\label{fig:crc_example_war}
	\end{subfigure}
	\caption{Code example demonstrating checkpointing gives rise to data inconsistency.}\label{fig:code_demo_incosistency}
\end{wrapfigure}
%\end{figure}

@@

%\textcolor{red}{reference work by Intel~\cite{baghsorkhi_cgo_2018}}

%Energy-harvesting devices execute software according to the {\em intermittent execution model}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on an energy-harvesting device executes {\em intermittently}
because energy sources are not always available to harvest and buffer
sufficient operational energy. An intermittent execution is composed of
operating periods interspersed with power
failures~\cite{dino,chain,alpaca,ratchet}. The frequency of failures depends on
the size of the device's energy storage buffer: a larger buffer allows longer
operating periods. 
%Energy-harvesters provide input power orders of magnitude less than operating power, making recharging negligible during operation.
%Intermittent execution is different from continuous execution. 
A power failure clears volatile state (e.g., registers and SRAM) while
non-volatile memory (e.g., FRAM) persists. Upon a power failure, control flows
to a prior point in the execution: by default, to the beginning of {\tt
main()}. Early intermittent systems preserved progress by periodically
checkpointing volatile execution context to non-volatile
memory~\cite{mementos}, sometimes using hardware
support~\cite{mottola2017harvos,hibernusplusplus,hibernus,idetic,quickrecall}.

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|}
		\hline
		Model & Data Copied to/from NVRAM \\
		\hline\hline
		Mementos~\cite{mementos}	& Registers + Stack     \\
		DINO~\cite{dino}	& Registers + Stack + WAR NV variables used in task\\
		Chain~\cite{chain}	& PC + NV variables used in task\\
		Alpaca~\cite{alpaca}	& PC + WAR NV variables used in task\\
		Ratchet~\cite{ratchet}, Clank~\cite{hicks_isca_2017} & Registers (requires NV main memory) \\
		Region Formation~\cite{baghsorkhi_cgo_2018} & Registers + Updated variables in task \\
		\hline
	\end{tabular}
	\caption{Non-volatile memory access for data consistency; \emph{PC}: program counter, \emph{WAR variables}: variables involved in WAR dependences, \emph{NV}: non-volatile.}
	\label{table:chechpoint_comparison}
\end{table}

@@

%Intermittent execution models incur overheads to checkpoint data~\cite{dino,ratchet,quickrecall,mementos}, manage channels~\cite{chain}, or privatize and commit data~\cite{alpaca}. Table~\ref{table:chechpoint_comparison} compares overheads for recent intermittent execution models (cf. \cite[Sec. 2.4]{alpaca}.) The mechanism responsible for overhead in each model varies, but the bulk of overhead in all models is a manipulating non-volatile memory. Checkpointing moves data to and from non-volatile memory, channel accesses manipulate non-volatile data~\cite{chain}, privatization copies from and commits to non-volatile memory~\cite{alpaca}, and idempotence solutions~\cite{ratchet} use {\em only} non-volatile memory.  
% We do not need to emphasize the following figure! (Sinan)
%\begin{figure}
%\begin{subfigure}[t]{.49\columnwidth}
%	\centering \includegraphics[width=\columnwidth]{figures/fram_write}
%	\caption{Cost of \emph{write} operation}
%\end{subfigure}%
%\begin{subfigure}[t]{.49\columnwidth}
%	\centering \includegraphics[width=\columnwidth]{figures/fram_read}
%	\caption{Cost of \emph{read} operation}
%\end{subfigure}
%	\caption{\textbf{Cost of accessing volatile (SRAM)/non-volatile (FRAM) memory during write/read operation.} \textcolor{red}{figures need to be merged into one}}\label{fig:framEnergy}
%\end{figure}
We measured the energy cost of volatile SRAM and non-volatile FRAM accesses on
the MSP430FR5969~\cite{msp430datasheet} MCU. We used EDB~\cite{edb} to monitor
voltage drop on a capacitor powering the device across 1600 read and write
accesses to each memory, randomizing accesses to avoid caching effects. Our
data showed that a non-volatile memory access consumes around 1.5 times the
energy of a volatile access. The data confirm that, because non-volatile memory
accesses are more expensive than volatile memory accesses, prior systems that
frequently access non-volatile memory may be artificially inefficient.
%~\cite{}. \todo{add citation to TI specification on FRAM energy cost}{Przemek}. 
% Figure~\ref{fig:framEnergy} summarizes the data, showing that a non-volatile memory access consumes around 1.5 times the energy of a volatile access in this device. This observation is also consistent with the TI specification~\cite{}. \todo{add citation to TI specification on FRAM energy cost}{Przemek} 
\sys is motivated by this observation: \sys 
moves data pages from FRAM to SRAM to maximize SRAM accesses and minimize FRAM
accesses, to decrease energy consumption and increase the time between power
failures.

>>>>

@@

\begin{wrapfigure}{t!}{0.5\textwidth}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example}
		\caption{Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration.}\label{fig:crc_example}
	\end{subfigure}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example_war}
		\caption{Execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 9) with stale reminder (line 3)}\label{fig:crc_example_war}
	\end{subfigure}
	\caption{Code example demonstrating checkpointing gives rise to data inconsistency.}\label{fig:code_demo_incosistency}
\end{wrapfigure}
%\end{figure}

Software running on an energy-harvesting device executes {\em intermittently}
because energy sources are not always available to harvest and buffer
sufficient operational energy. An intermittent execution is composed of
operating periods interspersed with power
failures~\cite{dino,chain,alpaca,ratchet}. The frequency of failures depends on
the size of the device's energy storage buffer: a larger buffer allows longer
operating periods. 
%Energy-harvesters provide input power orders of magnitude less than operating power, making recharging negligible during operation.
%Intermittent execution is different from continuous execution. 
A power failure clears volatile state (e.g., registers and SRAM) while
non-volatile memory (e.g., FRAM) persists. Upon a power failure, control flows
to a prior point in the execution: by default, to the beginning of {\tt
main()}. Early intermittent systems preserved progress by periodically
checkpointing volatile execution context to non-volatile
memory~\cite{mementos}, sometimes using hardware
support~\cite{mottola2017harvos,hibernusplusplus,hibernus,idetic,quickrecall}.

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|}
		\hline
		Model & Data Copied to/from NVRAM \\
		\hline\hline
		Mementos~\cite{mementos}	& Registers + Stack     \\
		DINO~\cite{dino}	& Registers + Stack + WAR NV variables used in task\\
		Chain~\cite{chain}	& PC + NV variables used in task\\
		Alpaca~\cite{alpaca}	& PC + WAR NV variables used in task\\
		Ratchet~\cite{ratchet}, Clank~\cite{hicks_isca_2017} & Registers (requires NV main memory) \\
		Region Formation~\cite{baghsorkhi_cgo_2018} & Registers + Updated variables in task \\
		\hline
	\end{tabular}
	\caption{Non-volatile memory access for data consistency; \emph{PC}: program counter, \emph{WAR variables}: variables involved in Write after Read (WAR) dependences, \emph{NV}: non-volatile.}
	\label{table:chechpoint_comparison}
\end{table}

@@

%Intermittent execution models incur overheads to checkpoint data~\cite{dino,ratchet,quickrecall,mementos}, manage channels~\cite{chain}, or privatize and commit data~\cite{alpaca}. Table~\ref{table:chechpoint_comparison} compares overheads for recent intermittent execution models (cf. \cite[Sec. 2.4]{alpaca}.) The mechanism responsible for overhead in each model varies, but the bulk of overhead in all models is a manipulating non-volatile memory. Checkpointing moves data to and from non-volatile memory, channel accesses manipulate non-volatile data~\cite{chain}, privatization copies from and commits to non-volatile memory~\cite{alpaca}, and idempotence solutions~\cite{ratchet} use {\em only} non-volatile memory.  
% We do not need to emphasize the following figure! (Sinan)
%\begin{figure}
%\begin{subfigure}[t]{.49\columnwidth}
%	\centering \includegraphics[width=\columnwidth]{figures/fram_write}
%	\caption{Cost of \emph{write} operation}
%\end{subfigure}%
%\begin{subfigure}[t]{.49\columnwidth}
%	\centering \includegraphics[width=\columnwidth]{figures/fram_read}
%	\caption{Cost of \emph{read} operation}
%\end{subfigure}
%	\caption{\textbf{Cost of accessing volatile (SRAM)/non-volatile (FRAM) memory during write/read operation.} \textcolor{red}{figures need to be merged into one}}\label{fig:framEnergy}
%\end{figure}
As an illustration, we measured the energy cost of volatile SRAM and non-volatile FRAM accesses on
the MSP430FR5969~\cite{msp430datasheet} MCU. We used EDB~\cite{edb} to monitor
voltage drop on a capacitor powering the device across 1600 read and write
accesses to each memory, randomizing accesses to avoid caching effects. Our
data showed that a non-volatile memory access consumes around 1.5 times the
energy of a volatile access. The data confirm that, because non-volatile memory
accesses are more expensive than volatile memory accesses, prior systems that
frequently access non-volatile memory may be artificially inefficient.
%~\cite{}. \todo{add citation to TI specification on FRAM energy cost}{Przemek}. 
% Figure~\ref{fig:framEnergy} summarizes the data, showing that a non-volatile memory access consumes around 1.5 times the energy of a volatile access in this device. This observation is also consistent with the TI specification~\cite{}. \todo{add citation to TI specification on FRAM energy cost}{Przemek} 
\sys is motivated by this observation: \sys 
moves data pages from FRAM to SRAM to maximize SRAM accesses and minimize FRAM
accesses, to decrease energy consumption and increase the time between power
failures.
